Towards Practical First-Order Model Counting

### 2024-08-14 doc/paper/paper.tex
minor edit

In this section, we present experimental results that analyse the scalability of
\Cranetwo{} in comparison to two other notable FOMC algorithms:
\textsc{FastWFOMC} and \textsc{ForcLift}. Our experiments involve two versions
of \Cranetwo{}: \Cranegreedy{} and \Cranebfs{}. Like its predecessor,
\Cranetwo{} has two modes for applying compilation rules to formulas: one that
uses a greedy search algorithm similar to \textsc{ForcLift} and another that
combines greedy and breadth-first search.

@@

We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\begin{gather*}
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land{}\\
  (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\end{gather*}
or, equivalently, in conjunctive normal form as
\begin{gather*}
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land{}\\
  (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\end{gather*}
Finally, we include the bijection-counting problem previously utilised by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Its
formulation in \FO{} is described in \cref{example:overall}. The equivalent
formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

@@

Since \textsc{FastWFOMC} does not support many-sorted logic, our experiments are
limited to formulas with a single domain. However, many of the counting problems
used in the experimental evaluation of
\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} become equivalent (in terms of
generating equivalent integer sequences) when restricted to a single domain.
Additionally, comparing \Cranetwo{} and \textsc{FastWFOMC} on a broader set of
problems is challenging because each problem needs to be represented in two
(quite different) logics: \FO{} and \UFO{}.

Nevertheless, the three benchmarks cover a wide range of possibilities. The
`friends' benchmark stands out as it uses multiple predicates and can be
expressed in \FO{} using just two variables without cardinality constraints or
counting quantifiers. The `functions' benchmark, on the other hand, can still be
handled by all the algorithms, but it requires cardinality constraints, counting
quantifiers, or more than two variables. Lastly, the `bijections' benchmark is
an example of a formula that \textsc{FastWFOMC} can handle but \textsc{ForcLift}
cannot.

@@

A considerable difference between \textsc{ForcLift} and the other two algorithms
is that \textsc{ForcLift} does not support arbitrary precision whereas the other
algorithms use the GNU Multiple Precision Arithmetic Library. Thus, when the
count exceeds finite precision, \textsc{ForcLift} returns $\infty$. For
evaluation purposes, we ran each algorithm on each benchmark using domains of
sizes $2^{1}, 2^{2}, 2^{3}$, and so on, until a timeout occurred after
\SI{1}{\hour}, the algorithm ran out of memory, or \textsc{ForcLift} returned
$\infty$. We separately measured compilation and inference time but primarily
focused on total runtime, dominated by the latter.

@@

\Cref{fig:plot} presents a summary of the experimental results. Only
\textsc{FastWFOMC} and \Cranebfs{} could handle the bijection-counting problem.
For this benchmark, the largest domain sizes these algorithms could accommodate
were \num{64} and \num{4096}, respectively. On the other two benchmarks,
\textsc{ForcLift} had the lowest runtime. However, due to its finite precision,
it only scaled up to domain sizes of \num{16} and \num{128} for `friends' and
`functions', respectively. \textsc{FastWFOMC} outperformed \textsc{ForcLift} in
the case of `friends', but not `functions', as it could handle domains of size
\num{1024} and \num{64}, respectively. Furthermore, both \Cranebfs{} and
\Cranegreedy{} performed similarly on both benchmarks. Similarly to the
`bijections' benchmark, \Cranetwo{} significantly outperformed the other two
algorithms, scaling up to domains of size \num{8192} and \num{67108864},
respectively.

>>>>

Our empirical evaluation sought to compare the runtime performance of {\Cranetwo} with the current state of the art, namely 
\textsc{FastWFOMC} and \textsc{ForcLift}. It is worth remarking that  \textsc{ForcLift} does not support arbitrary precision, and returns error for cases that requires arbitrary precision reasoning. 
Our experiments involve two versions
of \Cranetwo{}: \Cranegreedy{} and \Cranebfs{}. Like its predecessor,
\Cranetwo{} has two modes for applying compilation rules to formulas: one that
uses a greedy search algorithm similar to \textsc{ForcLift} and another that
combines greedy and breadth-first search.

@@

\subsection{Benchmarks}
We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\begin{gather*}
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land{}\\
  (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\end{gather*}
or, equivalently, in conjunctive normal form as
\begin{gather*}
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land{}\\
  (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\end{gather*}
Finally, we include the bijection-counting problem previously utilised by
\citeauthor{DBLP:conf/kr/DilkasB23}~\shortcite{DBLP:conf/kr/DilkasB23}. Its
formulation in \FO{} is described in \cref{example:overall}. The equivalent
formula in \Ctwo{} is
\begin{align*}
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta&\text{. }P(x, y))\land{}\\
  (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta&\text{. }P(x, y)).
\end{align*}
Similarly, in \UFO{} the same formula can be written as
\begin{gather*}
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y))\land{}\\
  (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x))\land{}\\
  (|P| = |\Delta|),
\end{gather*}
where $w^{-}(R) = w^{-}(S) = -1$.

@@

%Since \textsc{FastWFOMC} does not support many-sorted logic, our experiments are
%limited to formulas with a single domain. However, many of the counting problems
%used in the experimental evaluation of
%\textsc{Crane}~\cite{DBLP:conf/kr/DilkasB23} become equivalent (in terms of
%generating equivalent integer sequences) when restricted to a single domain.
%Additionally, comparing \Cranetwo{} and \textsc{FastWFOMC} on a broader set of
%problems is challenging because each problem needs to be represented in two
%(quite different) logics: \FO{} and \UFO{}.

The three benchmark families cover a wide range of possibilities. The
`friends' benchmark stands out as it uses multiple predicates and can be
expressed in \FO{} using just two variables without cardinality constraints or
counting quantifiers. The `functions' benchmark, on the other hand, can still be
handled by all the algorithms, but it requires cardinality constraints, counting
quantifiers, or more than two variables. Lastly, the `bijections' benchmark is
an example of a formula that \textsc{FastWFOMC} can handle but \textsc{ForcLift}
cannot.

@@

%A considerable difference between \textsc{ForcLift} and the other two algorithms
%is that \textsc{ForcLift} does not support arbitrary precision whereas the other
%algorithms use the GNU Multiple Precision Arithmetic Library. Thus, when the
%count exceeds finite precision, \textsc{ForcLift} returns $\infty$.

For
evaluation purposes, we ran each algorithm on each benchmark using domains of
sizes $2^{1}, 2^{2}, 2^{3}$, and so on, until an algorithm failed to handle a domain size due to timeout, out of memory error, or out of precision errors. While we separately measured compilation and inference time but primarily
focused on total runtime, dominated by the latter.

@@

\subsection{Results}
\Cref{fig:plot} presents a summary of the experimental results. Only
\textsc{FastWFOMC} and \Cranebfs{} could handle the bijection-counting problem.
For this benchmark, the largest domain sizes these algorithms could accommodate
were \num{64} and \num{4096}, respectively. On the other two benchmarks,
\textsc{ForcLift} had the lowest runtime. However, due to its finite precision,
it only scaled up to domain sizes of \num{16} and \num{128} for `friends' and
`functions', respectively. \textsc{FastWFOMC} outperformed \textsc{ForcLift} in
the case of `friends', but not `functions', as it could handle domains of size
\num{1024} and \num{64}, respectively. Furthermore, both \Cranebfs{} and
\Cranegreedy{} performed similarly on both benchmarks. Similarly to the
`bijections' benchmark, \Cranetwo{} significantly outperformed the other two
algorithms, scaling up to domains of size \num{8192} and \num{67108864},
respectively.
